---
title: "R-Ladies Cape Town: What's in a date?"
author: "Megan Beckett"
date: "16 August 2018"
output:
  html_notebook:
    fig_height: 4.5
    fig_width: 7
    toc: yes
---
# Overview
Date-time data can be frustrating to work with in R. R commands for date-times are generally unintuitive and change depending on the type of date-time object being used. Moreover, the methods we use with date-times must be robust to time zones, leap days, daylight savings times, and other time related quirks, and R lacks these capabilities in some situations. Lubridate makes it easier to do the things R does with date-times and possible to do the things R does not.

# Resources
- Make sure you check out the very handy R-Studio cheatsheet: https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf
- The chapter in R for Data Science on dates and times: http://r4ds.had.co.nz/dates-and-times.html

# Lubridate tutorial
Let's explore some of the functions within the lubridate package.

## But first, installation!
```{r}
# The easiest way to get lubridate is to install the whole tidyverse:
install.packages("tidyverse")

# Alternatively, install just lubridate:
install.packages("lubridate")

# Or the the development version from GitHub:
# install.packages("devtools")
devtools::install_github("tidyverse/lubridate")

# Load pacage
```

## Load package
```{r}
library(lubridate)
```


# Case study: Heatmap of running activities
# Read in data

```{r}
activities = read.csv("data/Activities.csv", stringsAsFactors = FALSE)
head(activities)
str(activities)
summary(activities)
```


