---
title: "Home Credit Default Risk"
author: "Megan Beckett"
date: "16 July 2018"
output:
  html_notebook:
    code_folding: hide
    fig_height: 4.5
    fig_width: 7
    number_sections: yes
    toc: yes
    css: ../includes/summarytools/summarytools.css
---
# Introduction
The purpose of this notebook is to show my approach, exploration and findings for the Kaggle competition being run by Home Credit with the question:

- **Can you predict how capable each applicant is of repaying a loan?**

There are several datasets provided, but in this analysis I have focused on the main application dataset, with some exploration of the Bureau dataset. 

This notebook includes my exploratory data analysis (step 00) and the subsequent steps are in R_scripts, with my general approach being:

- **00_explore:** Exploratory data analysis
- **01_preprocess:** Steps I took to clean and preprocess the data
- **02_model:** Training the model (I've left a few iterations in to show how it evolved)
- **03_test:** Evaluate model, look at importance of variables and make predictions on test dataset
- **04_submit:** Code for creating Kaggle submission file

# Set up
## Load packages required
Other packages required in subsequent steps are at the top of the relevant R_script.

```{r, message=FALSE,warning=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(skimr)
library(DT)
library(summarytools)
```

## Read in datasets
These are app_train, app_test and bureau.

```{r, echo=TRUE, results='hide'}
app_train <- read_csv("../raw_data/application_train.csv")
app_test <- read_csv("../raw_data/application_test.csv")
bureau <- read_csv("../raw_data/bureau.csv")
app_prev <- read_csv("../raw_data/previous_application.csv")
```

# Overview of dataset

## Summaries
Notes:

- I normally use a combination of str(), summary(), glimpse() to do this and recently explored skim().
- For **skim()**, I like how it splits the summary statistics across variable type and shows a basic histogram for numerical types.
- Missing values are also shown and this, along with the histograms, gives some indication of what to explore further.
- Subsequent to this, I have now used the package **summarytools** and specifically, dfSummary(). The histogram plots are even better (although render quite small) and the developers are working to improve the html output to work with Rmarkdown. 
- See this vignette: https://cran.r-project.org/web/packages/summarytools/vignettes/Recommendations-rmarkdown.html

```{r}
#str(app_train)
#summary(app_train)
#glimpse(app_train)
#skim(app_train)

```

View data using dfSummary() from summarytools package.

```{r, results = 'asis', message=FALSE, warning=FALSE}
view(dfSummary(app_train), method = 'render')

```

## Data types of variables
```{r}
table(sapply(app_train, class))
```

Notes: 

- There are 16 categorical (character) variables to convert to factors. 
- If I'd used read.csv, they could be converted automatically to factors, but read_csv is much faster.

# Exploring missing values
## Total missing values

```{r}
# Total missing values
total_missing = sum(is.na(app_train))
print(c("Total missing values in application train dataset:", total_missing))
```

## Missing values per variable

```{r}
# Absolute missing values per column
missing_val_abs <- colSums(is.na(app_train))
# Percentage missing values per column
missing_val_per <- missing_val_abs/nrow(app_train)*100
# Get column names
colnames <- colnames(app_train)
# Create dataframe of missing values and display as datatable
missing_summary = subset(data.frame(missing_val_abs, missing_val_per), missing_val_per>0) 
missing_summary %>% datatable(filter = 'top', colnames = c('Total missing values', 'Percentage missing values'), 
            options = list(pageLength = 15, autoWidth = F, order = list(2, 'desc')))
```


Notes: 

- There are 67 variables in total with missing values. 
- 49 variables have over 40% missing values.
- A large portion of these are to do with the decription of the typeof property.

# Exploration of single variables

## TARGET classification

```{r}
# Target variable - classification
app_train %>%  
  count(TARGET) %>%
  arrange(desc(n)) %>%
  ggplot(aes(reorder(TARGET, -n, FUN = min), n, fill = TARGET)) +
  geom_col() + 
  theme(legend.position = "none") + 
  labs(x = 'TARGET (0 = repaid, 1 = did not repay)',y = 'Count') +
  ggtitle("Distribution of classification")
```

Notes: 

- Binary classification (0 if loan repaid, 1 if loan not repaid). 
- There are many more loans paid on time than there aren't.


## Quantitative variable exploration {.tabset .tabset-fade .tabset-pills}

Notes: 

- All of the numerical variables shown here are right skewed. 
- This is can be dealt with in preprocessing, but I have left them as is for now. 

### Distribution of amount of credit 

```{r}
ggplot(app_train, aes(AMT_CREDIT)) + 
  geom_histogram(bins = 50) +
  ggtitle("Distribution of credit amount of loan")
```

### Distribution of loan annuity 

```{r}
ggplot(app_train, aes(AMT_ANNUITY)) + 
  geom_histogram(bins = 50) +
  ggtitle("Distribution of loan annuity amount")
```

### Distribution of goods price

```{r}
ggplot(app_train, aes(AMT_GOODS_PRICE)) + 
  geom_histogram(bins = 50) +
  ggtitle("Distribution of the price of goods amount for which loan is given")
```

### Income distribution

```{r}
ggplot(app_train, aes(AMT_INCOME_TOTAL)) + 
  geom_histogram(bins= 100) +
  xlim(0,1000000) +
  ggtitle("Distribution of income of applicant")
```

Summary of AMT_INCOME_TOTAL for app_train and app_test, respectively:

```{r}
summary(app_train$AMT_INCOME_TOTAL)
summary(app_test$AMT_INCOME_TOTAL)
```

Notes: There is a really large outlier of 117000000 for total income. It seems unlikely that such an individual would be applying for a loan and this could be a mistake. Consider replacing with NA during data cleaning.

## Categorical variable exploration {.tabset .tabset-fade .tabset-pills}

### Type of loan

```{r}
app_train %>%  
  count(NAME_CONTRACT_TYPE) %>%
  arrange(desc(n)) %>%
  ggplot(aes(reorder(NAME_CONTRACT_TYPE, -n, FUN = min), n, fill = NAME_CONTRACT_TYPE)) +
  geom_col() + 
  theme(legend.position = "none") + 
  labs(x = 'NAME_CONTRACT_TYPE',y = 'Count') +
  ggtitle("Contract type of loan applied for")
```

Note: Predominant loan type is cash loan.

### Gender

```{r}
app_train %>%  
  count(CODE_GENDER) %>%
  arrange(desc(n)) %>%
  ggplot(aes(reorder(CODE_GENDER, -n, FUN = min), n, fill = CODE_GENDER)) +
  geom_col() + 
  theme(legend.position = "none") + 
  labs(x = 'CODE_GENDER',y = 'Count') +
  ggtitle("No of loan applications per gender")
```

Note: More females applying for loans than males.

### Own a car

```{r}
app_train %>%  
  count(FLAG_OWN_CAR) %>%
  arrange(desc(n)) %>%
  ggplot(aes(reorder(FLAG_OWN_CAR, -n, FUN = min), n, fill = FLAG_OWN_CAR)) +
  geom_col() + 
  theme(legend.position = "none") + 
  labs(x = 'FLAG_OWN_CAR',y = 'Count') +
  ggtitle("No of people who own a car")
```

Note: About 50% less people own a car than those who don't

### Own any property

```{r}
app_train %>%  
  count(FLAG_OWN_REALTY) %>%
  arrange(desc(n)) %>%
  ggplot(aes(reorder(FLAG_OWN_REALTY, -n, FUN = min), n, fill = FLAG_OWN_REALTY)) +
  geom_col() + 
  theme(legend.position = "none") + 
  labs(x = '',y = 'Count') +
  ggtitle("No of people who own property")
```

Note: Unlike owning a car, there are many more who own property

### Level of education of loan applicants

```{r}
app_train %>%  
  count(NAME_EDUCATION_TYPE) %>%
  arrange(desc(n)) %>%
  ggplot(aes(reorder(NAME_EDUCATION_TYPE, -n, FUN = min), n, fill = NAME_EDUCATION_TYPE)) +
  geom_col() + 
  theme(legend.position = "none") + 
  labs(x = '',y = 'Count') +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  ggtitle("Level of education")
```

Notes: The majority have completed secondary school. There are a few with incomplete highschool or only lower secondary. This is an interesting flag.

### Income level of applicants

```{r}
app_train %>%  
  count(NAME_INCOME_TYPE) %>%
  arrange(desc(n)) %>%
  ggplot(aes(reorder(NAME_INCOME_TYPE, -n, FUN = min), n, fill = NAME_INCOME_TYPE)) +
  geom_col() + 
  theme(legend.position = "none") + 
  labs(x = '',y = 'Count') +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  ggtitle("Income level type")
```

Notes: The majority are working. There is a significant portion of pensioners, which is interesting. There are also some unemployed, students and those on materynity leave.
  
# Exploration of variables measuring time/age
## Age of clients
```{r}
summary(app_train$DAYS_BIRTH)
```
Notes: 

- Interesting, these are negative numbers. 
- According to Home Credit, these numbers indicate the relative age of person in number of days backwards from loan application date. 

Convert numbers to absolute value and in years to view a meaningful plot.

```{r}
age = as.data.frame(app_train$DAYS_BIRTH/(-365))
colnames(age) = 'AGE_YEARS'
summary(age)
ggplot(age, aes(AGE_YEARS)) + 
  geom_histogram(bins = 50) +
  ggtitle("Distribution of age of loan applicants")
```

Notes: Median age is in the 40's and there is a wide distribution in ages from 20 to 70 years old.

# Days of employment
```{r}
summary(app_train$DAYS_EMPLOYED)
```

```{r}
ggplot(app_train, aes(DAYS_EMPLOYED)) + 
  geom_histogram(bins = 50) +
  ggtitle("Distribution of days employed")
```

Notes: 

- As with age in number of days, days of emplyment is no days before the application the person started current employment. 
- Median is -1213 which is about 3 years prior to application.
- But, positive numbers don't make sense and there are massive outliers equating to about 1000 years (365243 days)

Create subset of these outliers to investigate.

```{r}
employed_outliers <- subset(app_train, DAYS_EMPLOYED > 0)
employed_normal <- subset(app_train, DAYS_EMPLOYED <= 0)
print("Summary of outliers TARGET")
summary(employed_outliers$TARGET)
print("Summary of normal TARGET")
summary(employed_normal$TARGET)
```


Notes: 

- The outliers (all have same value of 365243 days) only default on 5.4% of loans whereas those with reasonable employment records, don't repay 8.66% of time.
- Investigate later by replacing these values with NA before imputation to replace missing values and see if has an effect.

# A quick look at additional datasets

## Bureau

```{r}
summary(bureau)
skim(bureau)
```

What are the unique values for the variable CREDIT_ACTIVE?

```{r}
unique(bureau$CREDIT_ACTIVE)
```

Notes:

- After first doing a logistic regression using only the applications data set, I am going to see what features I can create from the Bureau data set by grouping according to the loan id and then calculating some sums of numerical values of interest.
- I think the interesting ones to look at are:
- CREDIT_DAY_OVERDUE: Number of days past due on CB credit at the time of application for related loan in our sample
- AMT_CREDIT_SUM: Current credit amount for the Credit Bureau credit
- AMT_CREDIT_SUM_DEBT: Current debt on Credit Bureau credit
- AMT_CREDIT_SUM_OVERDUE: Current amount overdue on Credit Bureau credit


## Previous application

```{r}
summary(app_prev)
```

What are the unique values for the variables CODE_REJECT_REASON and NAME_CONTRACT_STATUS, respectively?

```{r}
unique(app_prev$CODE_REJECT_REASON)
unique(app_prev$NAME_CONTRACT_STATUS)
```











